\textsc{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}

### 2016-07-29 redaction/dhp_draft_july2016.tex
Petites modifications

Following \citet{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis). However, many common empirical facts support the existence of additive costs. As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. The Alchian and Allen conjecture (\citealp{alchian}), points out that the relative price of two varieties of some good will depend on the level of trade costs, does rely on the existence of additive costs: The relative demand for more expensive/higher quality product goods should increase with trade cost (``shipping the good apples out'').
\citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of his results though. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fab unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price is estimated to be well below unity. Also, their estimates imply that doubling freight costs increases average fab export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

The non-linearity of Equation (\ref{eq:estimatedequation}) implies that it cannot be estimated using standard linear estimators. Therefore, all estimates are performed using non-linear least squares. The basis of the method is to approximate the model by a linear one and to refine the parameters by successive iterations. The intuitive criterion for convergence is that the sum of squares does not decrease from one iteration to the next.
% In our case, due to computational limitations implied by the size of our dataset, we implement 100 iterations and set the convergence criterion for successive parameter estimates and for the residual sum of squares at 0.01.
%GD : cela me semble être de la cuisine sans grand intérêt.
Finally, to eliminate the potential influence of outliers, we excluded 5 percent of the upper and lower tails of the distribution in the regression variables
%%and all our three measures of trade costs are bounded by 0 as minimal value.
%%Pas compris. Parfois un, non ?
These cut-offs are aimed at eliminating reporting or coding errors.\medskip

@@

At this stage, two interesting comments can be made from Figure \ref{fig:decomp_TC_3d}. First, the magnitude of transport costs is higher for ocean shipping than for air transport. This result also holds considering ad-valorem costs alone (Figure \ref{fig:mult_alone_withadd}). Second, in both transport modes, the additive component appears of sizeable importance: As mean value over the period, it amounts to 48.2\% of total transport costs in Vessel, and 42.3\% in Air. We go on investigating this result further in the next subsection.\footnote{Figure \ref{fig:decomp_TC_3d} also delivers interesting results with respect to the time trends of transport costs. We come back to this aspect in Section \ref{sec:results_trends}.}

@@

In Figure \ref{fig:TC_addplusmult_compeffects_excl}, we report the sum of the ad-valorem and the additive costs (re-scaled to start from the reference value 100), for each transport mode.\footnote{For each mode, and by year, we build the total value of transport costs as the weighted sum of both the ad valorem index and the additive index, taking as weighting scheme the relative share of each type of costs in the overall total cost in 1974 (using the figures at the root of Figure \ref{fig:mult_alone_withadd}), i.e. respectively 48.2 \% for additive costs in Vessel and 42?3\% for Air in 1974.}

@@

This paper empirically studies the magnitude of additive (or per-kg) costs in international transport costs, by exploiting the differences between the import and the export prices. Using SITC 3 and 4- digit cif-fab unit values taken from the US import database over 1974-2013, we estimate the two components of transport costs, by transport mode (air or ocean).  Our results may be summarized in three main findings. First, we obtain that additive costs are 2.8\% of fab unit values for ocean shipping (and ad-valorem ones 3.2\%). These values are respectively equal to 1.8\% and 2.5\% for air transport. Second, we show that taking additive costs into account improves the fit of the modelling of transport costs. All goodness-of-fit measures point out to this conclusion, which holds for both transport modes and all years considered. Third, we also use the time dimension of our data to characterize the evolution of transport costs. After correcting for composition effects, we find that all types of transport costs have been roughly constant from 1974 to 1984 and then steadily decreased by 40\% over the period 1984-2013. Yet, this steady decline hides shifts in the relative importance of additive and ad-valorem. While most of the early decline in air transport costs can be explained by the ad-valorem component, this component nearly doubled in the 2000s.
Further, the inclusion of additive costs yields to the conclusion the decrease of overall trade costs are decreased in air transport and ocean shipping roughly the same path and the same magnitude of order. This last result stands in contrast with related studies (\citealp{hummels2007} (that omit the role of additive costs), which rather obtain a larger decrease in air transport costs, relative to ocean shipping costs. \citealp{Behar_Venables}). \todo{LP 28-07-2016: En reponse aux doutes de GD sur ce point, est-ce plus convaincant ainsi?} In all three aspects, our results point the importance of the additive component in accounting for international transport costs.

>>>>

Following \citet{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis). However, many common empirical facts support the existence of additive costs. As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. The Alchian and Allen conjecture (\citealp{alchian}), points out that the relative price of two varieties of some good will depend on the level of trade costs, does rely on the existence of additive costs: The relative demand for more expensive/higher quality product goods should increase with trade cost (``shipping the good apples out'').
\citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of his results though. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international costs. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fab unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price is estimated to be well below unity. Also, their estimates imply that doubling freight costs increases average fab export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

The non-linearity of Equation (\ref{eq:estimatedequation}) implies that it cannot be estimated using standard linear estimators. Therefore, all estimates are performed using non-linear least squares. The basis of the method is to approximate the model by a linear one and to refine the parameters by successive iterations. The intuitive criterion for convergence is that the sum of squares does not decrease from one iteration to the next.
Finally, to eliminate the potential influence of outliers, we excluded 5 percent of the upper and lower tails of the distribution in the regression variables.
These cut-offs are aimed at eliminating reporting or coding errors.\medskip

@@

At this stage, two interesting comments can be made from Figure \ref{fig:decomp_TC_3d}. First, the magnitude of transport costs is higher for ocean shipping than for air transport. This result also holds considering ad-valorem costs alone (Figure \ref{fig:mult_alone_withadd}). Second, in both transport modes, the additive component appears of sizeable importance: as mean value over the period, it amounts to 48.2\% of total transport costs in Vessel, and 42.3\% in Air. We go on investigating this result further in the next subsection.\footnote{Figure \ref{fig:decomp_TC_3d} also delivers interesting results with respect to the time trends of transport costs. We come back to this aspect in Section \ref{sec:results_trends}.}

@@

In Figure \ref{fig:TC_addplusmult_compeffects_excl}, we report the sum of the ad-valorem and the additive costs (re-scaled to start from the reference value 100), for each transport mode.\footnote{For each mode, and by year, we build the total value of transport costs as the weighted sum of both the ad valorem index and the additive index, taking as weighting scheme the relative share of each type of costs in the overall total cost in 1974 (using the figures at the root of Figure \ref{fig:mult_alone_withadd}), i.e. respectively 48.2 \% for additive costs in Vessel and 42.3\% for Air in 1974.}

@@

This paper empirically studies the magnitude of additive (or per-kg) costs in international transport costs, by exploiting the differences between the import and the export prices. Using SITC 3 and 4- digit cif-fab unit values taken from the US import database over 1974-2013, we estimate the two components of transport costs, by transport mode (air or ocean).  Our results may be summarized in three main findings. First, we obtain that additive costs are 2.8\% of fab unit values for ocean shipping (and ad-valorem ones 3.2\%). These values are respectively equal to 1.8\% and 2.5\% for air transport. Second, we show that taking additive costs into account improves the fit of the modelling of transport costs. All goodness-of-fit measures point out to this conclusion, which holds for both transport modes and all years considered. Third, we also use the time dimension of our data to characterize the evolution of transport costs. After correcting for composition effects, we find that all types of transport costs have been roughly constant from 1974 to 1984 and then steadily decreased by 40\% over the period 1984-2013. Yet, this steady decline hides shifts in the relative importance of additive and ad-valorem. While most of the early decline in air transport costs can be explained by the ad-valorem component, this component nearly doubled in the 2000s.
Further, the inclusion of additive costs yields to the conclusion the decrease of overall trade costs are decreased in air transport and ocean shipping roughly the same path and the same magnitude of order. This last result stands in contrast with related studies (\citealp{hummels2007}) (that omit the role of additive costs), which rather obtain a larger decrease in air transport costs, relative to ocean shipping costs (\citealp{Behar_Venables}). In all three aspects, our results point the importance of the additive component in accounting for international transport costs.
